Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-15 aistats.tex

For exponential families, MAP perfectly fits into this framework, as
\aligns{
	f(\theta) = A(\theta) - \expect{\lin{T(X), \theta}}
}
is $1$-smooth and $1$-strongly convex relative to $A$.
Our goal is then to find an applicable convergence rate for SMD under relative smoothness.
Several pieces of work address this question under various assumptions on stochasticity \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}, 
but they have so far lacked concrete examples.
In the next section, we challenge their hypothesis thanks to the MAP.

@@

To analyze stochastic algorithms, one also needs to quantify the randomness.
While many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}, only a few have been adapted to SMD with relative smoothness.
We review these in the light of  the MAP and provide a summary in \cref{tbl:assumptions}.

@@

\subsubsection{Bounded Variance}
\citet{hanzely2018fastest} assume the covariance between negative gradient and next iterate is upper bounded at every time step: $\forall n \geq 1, \forall \hat \nat_{n-1},$
\alignn{
	\Cov(-g_n(\hat \nat_{n-1}) , \MAPt) \leq \lr_n C
	\label{eq:hanzely}
}
for some constant $C$.
For a Gaussian with known variance ($A(\theta) = \frac{1}{2}\norm{\theta}^2$),
this definition recovers the variance of the stochastic gradient
\[
	\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2}\leq C.
\]
More generally, if $A$ is $\stgcvx$-strongly convex w.r.t. $\norm{\cdot}$,
the LHS of \cref{eq:hanzely} is bounded by
$\expecti{\frac{1}{\stgcvx}\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2_*}$.
Then a bound on the variance of stochastic gradients is sufficient for \cref{eq:hanzely}
to hold.
%Another perspective on this covariance is as the expectation of the symmetrized Bregman $\cS_\conj(\m_1, \m_2) = \bregmanconj(\m_1, \m_2) + \bregmanconj(\m_2, \m_1)$ between stochastic and deterministic updates
%\aligns{
%\Cov(-g_n(\hat \nat_{n-1}) , \MAPt) = \E[\cS(\MAPm - \lr g_n  )]
%}
%thus showing positivity.
Under this assumption, \citet[Lem.4.8]{hanzely2018fastest} prove a $O(1/n)$ convergence rate with $O(1/n)$ step-sizes,
by applying tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
However, in general, this assumption rarely holds.
For the MAP, this translates into $\forall n \geq 1,$
\alignn{
\expect[T_n]{\lin{T_n - \meanp^* , \MAPt}} \leq \lr_n C
}
which fails to hold on the gaussian variance example, for instance. Indeed, we cannot control the samples $x_i$, therefore there is a non-zero probability they make $\nabla\conj(\MAPt) = \hat\meanp_t$ arbitrarily close to zero. Since the conjugate of the log partition is $\conj=-0.5\log$, this makes $\MAPt$ unbounded.

%  \TODO{ DAMIEN short proof or formalize.}

\subsubsection{Bounded Variance at the Minimum}

\citet{dragomir2021fast} use the weaker assumption
\alignn{
	\expect[\tilde g]{
		\cB_{A^*}(\MAPm - 2\lr g(\theta_*), \MAPm)
	} \leq 2 \lr^2 C \; ,
	\quad
	\forall n \geq 1.
	\label{eq:dragomir}
}
This bounds the variance of the gradients at the minimum with a metric depending on the iterate $\MAPm$. This bound should hold for every iterate we may encounter on the stochastic trajectory.
% As observed by Frederick, this is really weaker than Hanzely, because it is only the first half of their assumption. There is only a factor 2 that's preventing a direct inequality.
Under this assumption, they prove convergence up to a variance ball with a constant step size.
Their descent lemma \citep[Eq.(12)]{dragomir2021fast}
is the analog of modern analysis of SGD~\citep[Th.3.2]{gower2019sgd},
meaning we can plug in a $O(1/n)$ decreasing step-size and obtain a $O(1/n)$ convergence rate.

This seems to be a solution to our problem, but it suffers from the same issues as \citet{hanzely2018fastest}.
For barrier objectives, \cref{eq:dragomir} does not hold uniformly on every path of iterates,
because these paths can get arbitrarily close to boundaries where $\bregmanconj$ diverges.

@@

\subsection{Issues with current analyses}

\fdk{They do not hold for problems we care about}

\fdk{Existing techniques only deal with fixed step-sizes and show convergence to a ball}

>>>>

For exponential families, MAP perfectly fits into this framework, as
\aligns{
	f(\theta) = A(\theta) - \expect{\lin{T(X), \theta}}
}
is $1$-smooth and $1$-strongly convex relative to $A$.
Our goal is then to find an applicable convergence rate for SMD under relative smoothness.

@@

To analyze stochastic algorithms, one also need to quantify the randomness.
While many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}, only a few have been adapted to SMD with relative smoothness \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}, but they have so far been lacking concrete examples.
We review these analyses in the light of the MAP and provide a summary in \cref{tbl:assumptions}.

@@

%Before we proceed, let $f_X$ be a stochastic estimate of $f = \expecti{f_X}$, with stochastic gradients $g_X(\nat) =\nabla f_X(\nat)$. In our case $f_X(\nat) = - \log p(X\cond \nat)$ and $g_X(\nat) = \m - T(X)$.

\subsubsection{Analogs of the Variance}
let us introduce $\cS_\conj(\m_1, \m_2) = \bregmanconj(\m_1, \m_2) + \bregmanconj(\m_2, \m_1)$ th symmetrized Bregman induced by $\conj$.
\citet{hanzely2018fastest} assume that the expectation of $\cS$ between stochastic and deterministic updates verifies
\alignn{
\expect[g]{\cS_\conj \paren{
	\hat\m_{n} - \lr g(\hat \nat_{n}),
	\hat\m_{n}   -\lr \nabla f(\hat \nat_{n})
}} \leq \lr^2 C
\label{eq:hanzely}
}
for all possible iterates $\MAPt$, relevant step-sizes $\lr$ and for some constant $C$.
When $A(\theta) = \frac{1}{2}\norm{\theta}^2$ or when $A$ is $\stgcvx$-strongly convex,
this definition recovers the variance of the stochastic gradient
\[
	\expect[g\!\!]{\inv{\stgcvx}\norm{\nabla f(\theta) - g(\theta)}^2}\leq C.
\]
Under this assumption, \citet[Lem.4.8]{hanzely2018fastest} prove a $O(1/n)$ convergence rate on function values with $O(1/n)$ step-sizes and tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.

\citet{dragomir2021fast} define the weaker assumption
\alignn{
	\expect[\tilde g]{
		\cB_{A^*}(\MAPm - 2\lr g(\theta_*), \MAPm)
	} \leq 2 \lr^2 C \; .
	\label{eq:dragomir}
}
When $\logpart$ is $\stgcvx$-strongly-convex, we recover the variance of the gradients at the optimum
\[
	\expect[g\!\!]{\inv{\stgcvx}\norm{\nabla g(\theta^*)}^2}\leq C.
\]
Using their descent lemma \citep[Eq.(12)]{dragomir2021fast} with the $O(1/n)$ step-size used by \citet[Th.3.2]{gower2019sgd} for SGD, we obtain a $O(1/n)$ convergence rate, on the Bregman with \emph{reversed} arguments $\bregman(\nat^*, \hat\nat_n)$.

These two analyses seem promising for~\eqref{problem}, but none of these assumptions hold in front of barrier objectives such as the $-\log$ from \S\ref{ssec:gaussian-variance}.
Indeed, we cannot control the samples $x_i$, therefore there is a non-zero probability they make $\nabla\conj(\MAPt) = \hat\meanp_t$ arbitrarily close to zero. Since the conjugate of the log partition is $\conj=-0.5\log$, this makes $\MAPt$ unbounded.

@@
